---
title: "Coursera - Reproducible Research - Project 2"
output: html_notebook
---

Author: Tomy Bolanos
Date: 2019-09-22

## Exploring the NOAA Storm Database

### Assignment Brief

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

### Data processing 

```{r}

# Download the file if it does not exist.

if(!file.exists("repdata-data-StormData.csv.bz2")) {  
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                destfile = "repdata-data-StormData.csv.bz2", method = "curl")}

# Load data set
fileData <- read.csv(bzfile("repdata-data-StormData.csv.bz2"), sep=",", header = TRUE)
# Identifybg variables
names(fileData)
# Subsetting the storm data in the memory
dsStorms <- fileData[ , c('EVTYPE', 'FATALITIES', 'INJURIES', 'PROPDMG', 'PROPDMGEXP', 
                         'CROPDMG', 'CROPDMGEXP')]

# Convert H, K, M, B units to calculate Property Damage 
dsStorms$pd <- 0
dsStorms[dsStorms$PROPDMGEXP == "H", ]$pd <- 
  dsStorms[dsStorms$PROPDMGEXP == "H", ]$PROPDMG * 10^2

dsStorms[dsStorms$PROPDMGEXP == "K", ]$pd <- 
  dsStorms[dsStorms$PROPDMGEXP == "K", ]$PROPDMG * 10^3

dsStorms[dsStorms$PROPDMGEXP == "M", ]$pd <- 
  dsStorms[dsStorms$PROPDMGEXP == "M", ]$PROPDMG * 10^6

dsStorms[dsStorms$PROPDMGEXP == "B", ]$pd <- 
  dsStorms[dsStorms$PROPDMGEXP == "B", ]$PROPDMG * 10^9

# Converting the H, K, M, B into units to be able to calculate Crop Damage
dsStorms$cd <- 0

dsStorms[dsStorms$CROPDMGEXP == "H", ]$cd <- 
  dsStorms[dsStorms$CROPDMGEXP == "H", ]$CROPDMG * 10^2

dsStorms[dsStorms$CROPDMGEXP == "K", ]$cd <- 
  dsStorms[dsStorms$CROPDMGEXP == "K", ]$CROPDMG * 10^3

dsStorms[dsStorms$CROPDMGEXP == "M", ]$cd <- 
  dsStorms[dsStorms$CROPDMGEXP == "M", ]$CROPDMG * 10^6

dsStorms[dsStorms$CROPDMGEXP == "B", ]$cd <- 
  dsStorms[dsStorms$CROPDMGEXP == "B", ]$CROPDMG * 10^9

```

### Analysis and Results

#### 1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

```{r}
library(ggplot2)
# Plotting the Number of Fatalities By the Most Harmful Event Types
fat <- aggregate(FATALITIES ~ EVTYPE, data = dsStorms, sum)
fat <- fat[order(-fat$FATALITIES), ][1:20, ]
fat$EVTYPE <- factor(fat$EVTYPE, levels = fat$EVTYPE)

ggplot(fat, aes(x = EVTYPE, y = FATALITIES)) + 
  geom_bar(stat = "identity", fill = "red", las = 3) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("EVENTS") + ylab("FATALITIES") +
  ggtitle("Fatalities by Top 20 Weather Events")
```


### 2. Across the United States, which types of events have the greatest economic consequences ?

```{r}
# Plotting the Number of Damages Catagorized By the Most Harmful Event 
dam <- aggregate(pd + cd ~ EVTYPE, data = dsStorms, sum)
names(dam) <- c("EVTYPE", "TDAMAGE")
dam <- dam[order(-dam$TDAMAGE), ][1:20, ]
dam$EVTYPE <- factor(dam$EVTYPE, levels = dam$EVTYPE)

ggplot(dam, aes(x = EVTYPE, y = TDAMAGE)) + 
    geom_bar(stat = "identity", fill = "red", las = 3) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("EVENT") + ylab("DAMAGES (US$)") +
  ggtitle("Property & Crop Damages by Top 20  Events")
```

